---
title: "causens: an R package for causal sensitivity analysis methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{causens_user_guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Methods to estimate causal quantities often rely no unmeasured confounding. However, 
in practice, this is often not the case. `causens` is an R package that provides
methods to estimate causal quantities in the presence of unmeasured confounding. In
particular, `causens` implements methods from the following papers:

- Brumback, B. A., Hernán, M. A., Haneuse, S. J., & Robins, J. M. (2004). Sensitivity analyses for unmeasured confounding assuming a marginal structural model for repeated measures. Statistics in medicine, 23(5), 749-767.
- Li, L., Shen, C., Wu, A. C., & Li, X. (2011). Propensity score-based sensitivity analysis method for uncontrolled confounding. American journal of epidemiology, 174(3), 345-353.

# Installation

The package can be installed from GitHub via `devtools`.

```{r}
install.packages("devtools")
library(devtools)
devtools::install_github("Kuan-Liu-Lab/causens")
```

# Implemented Methods

## Frequentist Methods (Brumback et al. 2004, Li et al. 2011)

# Synthetic Data Example

We posit the following data generating process:

\begin{align*}
  X_1 &\sim \text{Normal}(0, 1) \\
  X_2 &\sim \text{Normal}(0, 1) \\
  X_3 &\sim \text{Normal}(0, 1) \\
  U &\sim \text{Bern}(0.5) \\
  Z &\sim \text{Bern}(\text{logit}^{-1}(\alpha_{uz} U + 0.25 X_1 + 0.5 X_2 - 0.5 X_3)) \\
  Y1 &\sim \text{Normal}(\beta_{uy} U + 0.5 X_1 + 0.5 X_2 - 0.5 X_3) \\
  Y0 &\sim \text{Normal}(0.5 X_1 + 0.5 X_2 - 0.5 X_3) \, .
\end{align*}

Using consistency, we define $Y = Y_1 Z + Y_0 (1 - Z)$. Note that, 
in `simData.R`, we provide some options to allow the simulation 
procedure to be more flexible. Parameters $\alpha_{uz}$ and $\beta_{uy}$
dictate how the unmeasured confounder $U$ affects the treatment $Z$ and
the outcome $Y$, respectively; by default, they are set to 0.2 and 0.5. 

```{r}
data <- simulate_data(N = 10000, seed = 123, alpha_uz = 1,
                      beta_uy = 1, treatment_effects = 1)
```

## Using [Frequentist Methods](## Frequentist Methods (Brumback et al. 2004, Li et al. 2011)

For these methods, we require to posit our own sensitivity function $c(z, e)$.
In practice, we don't know these values and they should be provided by subject
matter experts. However, for oracle models (or at least the one we posited), we
are able to retrieve the true sensitivity function, which is given by 
$c(z, e) \approx 0.24$ for both $z = 0, 1$ and for all $e \in (0, 1)$.

```{r}
# Posit incorrect treatment model since U is "missing"
trt_model <- glm(Z ~ X.1 + X.2 + X.3, data = data, family = binomial())

# Frequentist sensitivity analysis
causens(trt_model, data, "Z", "Y", method = "Li", c1 = 0.25, c0 = 0.25)
```
